---
title: "Untitled"
author: "Anous Ahmed"
format: pdf
editor: visual
---

#Question 1 - EDA

##Dataset Export Quality Data for Passenger Cars, 1963-2014


**Perform as many of the following tasks as you can. If you cannot perform a task, write a brief explanation of how you would perform it. The questions that are starred (\*) require only text responses.**

**1. Load the dataset and display the first few rows.**

```{r setup}
library("tidyverse")
library("stevedata")
```

```{r} 
eq_passengercars = eq_passengercars
```

```{r}
head(eq_passengercars, 10)
```

**2. Display the structure and summary of the dataset.**

```{r}
# check the structure
str(eq_passengercars)
```

```{r}
# check for rows and columns
summary(eq_passengercars)
```

```{r}
# check for rows and columns
dim(eq_passengercars)
```

**3. Is the dataset tidy? If not, what makes it messy? If you were to make it tidy, what would be the "unit of observation" for each row?**

The dataset is not tidy because the variables, **category** and **type** have multiple pieces of information that must be separated and inserted in different columns. The **category** column contains both a measurement type and and confidence interval descriptors. On the other hand, the **type** column contains information on numerical data, general categories and product types. Therefore, the dataset would need to be tidy in order to carry out further analysis. 

The **unit of observation** for each row would be the measurement of a country's export for each year and product type in a particular sector. 

**4. If you think the data is not tidy, tidy the dataset.**

```{r}
eq_passengercars_tidy = eq_passengercars %>%
  separate(type, into = c("industry_sector", "product_type"), sep = ",", remove = T) %>%
  separate(industry_sector, into = c("sector_code", "sector_name"), sep = "\\. ", remove = T) %>%
  separate(category, into = c("measurement", "interval"), sep = " - ", fill = "right")
```

```{r}
head(eq_passengercars_tidy, 5)
```

**5. Check for missing values and duplicates. If there are any, handle them appropriately.**

```{r}
missing_values = eq_passengercars_tidy %>%
    summarise_all(~ sum(is.na(.)))

print(missing_values)
```

```{r}
# calculating the proportion of missing values
missing_frac = eq_passengercars_tidy %>%
    summarise_all(~ sum(is.na(.)) / n())

print(missing_frac)
```

**6. Plot some of the variables in the dataset. What are some interesting trends you can see? (If plotting all countries is not possible, pick 6-8 countries of interest to you.)**

```{r}
selected_countries = c("Australia", "United States", "Japan", "United Kingdom", "Germany", "China")
```


```{r}
library(ggplot2)
library(dplyr)

dist_graph = eq_passengercars_tidy %>%
  filter(measurement == "Export Quality Index", country %in% selected_countries)

ggplot(dist_graph, aes(x = year, y = value, color = country)) +
  geom_line() +
  geom_point() +
  labs(title = "Export Quality Index Over Time",
       x = "Year",
       y = "Export Quality Index",
       color = "Country") +
  theme_minimal()
```


## Now, answer the following questions using the eq_passengercars dataset:

**1. Which country has the highest "Trade value of exports"? Which year was this?**

```{r}
highest_trade_exports = eq_passengercars_tidy %>%
# Filter the rows that correspond to "Trade value of exports"
  filter(measurement == "Trade value of exports") %>%
# filter the row where "value" is the maximum
  filter(value == max(value, na.rm = TRUE)) %>%
  select(country, year, value)
```

```{r}
# Germany had the highest Trade value of exports in 2014 
highest_trade_exports
```

**2. Which year had the lowest "Trade value of exports"? Which country was this?**

```{r}
lowest_trade_exports = eq_passengercars_tidy %>%
# Filter the rows that correspond to "Trade value of exports"
  filter(measurement == "Trade value of exports") %>%
# filter the row where "value" is the maximum
  filter(value == min(value, na.rm = TRUE)) %>%
  select(country, year, value)
```

```{r}
# South Africa had the highest Trade value of exports in 1964 
highest_trade_exports
```

**3. Which year did your country/any country of your interest have the highest "Export Quality Index"?**

```{r}
highest_export_index = eq_passengercars_tidy %>%
# Filter the rows that correspond to "Export Quality Index"
  filter(measurement == "Export Quality Index") %>%
# I'll be choosing the United States  
  filter(country == "United States") %>%
# filter the row where "value" is the maximum
  filter(value == max(value, na.rm = TRUE)) %>%
  select(country, year, value)
```

```{r}
# The United States had the highest Export Quality Index in 1984 
highest_export_index
```

**4. Which year, on average, had the greatest decrease in "Trade value of exports" from the previous year?**

```{r}
highest_trade_value = eq_passengercars_tidy %>%
# Filter the rows that correspond to "Trade value of exports"
  filter(measurement == "Trade value of exports") %>%
# Calculate the average export value per year
  group_by(year) %>%
  summarise(avg_trade_exports = mean(value, na.rm = TRUE)) %>%
# Ensure the data is ordered by year
  arrange(year) %>%
# Compute the drop from the previous year.
# Here, we calculate how much lower the current year's average is compared to the previous year's average.
  mutate(decrease = lag(avg_trade_exports) - avg_trade_exports) %>%
# Step 5: Find the year with the greatest drop (i.e. the maximum decrease value)
  filter(decrease == max(decrease, na.rm = TRUE)) %>%
  select(year, decrease)
```

```{r}
highest_trade_value
```
