---
title: "Assignment 2"
author: "Anous Ahmed"
format: pdf
editor: visual
---

#Question 1 - EDA

##Dataset Export Quality Data for Passenger Cars, 1963-2014

**Perform as many of the following tasks as you can. If you cannot perform a task, write a brief explanation of how you would perform it. The questions that are starred (\*) require only text responses.**

**1. Load the dataset and display the first few rows.**

```{r setup}
library("tidyverse")
library("stevedata")
```

```{r}
eq_passengercars = eq_passengercars
```

```{r}
head(eq_passengercars, 10)
```

**2. Display the structure and summary of the dataset.**

```{r}
# check the structure
str(eq_passengercars)
```

```{r}
# check for rows and columns
summary(eq_passengercars)
```

```{r}
# check for rows and columns
dim(eq_passengercars)
```

**3. Is the dataset tidy? If not, what makes it messy? If you were to make it tidy, what would be the "unit of observation" for each row?**

The dataset is not tidy because the variables, **category** and **type** have multiple pieces of information that must be separated and inserted in different columns. The **category** column contains both a measurement type and and confidence interval descriptors. On the other hand, the **type** column contains information on numerical data, general categories and product types. Therefore, the dataset would need to be tidy in order to carry out further analysis.

The **unit of observation** for each row would be the measurement of a country's export for each year and product type in a particular sector.

**4. If you think the data is not tidy, tidy the dataset.**

```{r}
eq_passengercars_tidy = eq_passengercars %>%
  separate(type, into = c("industry_sector", "product_type"), sep = ",", remove = T) %>%
  separate(industry_sector, into = c("sector_code", "sector_name"), sep = "\\. ", remove = T) %>%
  separate(category, into = c("measurement", "interval"), sep = " - ", fill = "right")
```

```{r}
head(eq_passengercars_tidy, 5)
```

**5. Check for missing values and duplicates. If there are any, handle them appropriately.**

```{r}
missing_values = eq_passengercars_tidy %>%
    summarise_all(~ sum(is.na(.)))

print(missing_values)
```

```{r}
# calculating the proportion of missing values
missing_frac = eq_passengercars_tidy %>%
    summarise_all(~ sum(is.na(.)) / n())

print(missing_frac)
```

**6. Plot some of the variables in the dataset. What are some interesting trends you can see? (If plotting all countries is not possible, pick 6-8 countries of interest to you.)**

```{r}
selected_countries = c("Australia", "United States", "Japan", "United Kingdom", "Germany", "China")
```

```{r}
library(ggplot2)
library(dplyr)
```

```{r}
dist_graph = eq_passengercars_tidy %>%
  filter(measurement == "Export Quality Index", country %in% selected_countries)

ggplot(dist_graph, aes(x = year, y = value, color = country)) +
  geom_line() +
  geom_point() +
  labs(title = "Export Quality Index Over Time",
       x = "Year",
       y = "Export Quality Index",
       color = "Country") +
  theme_minimal()
```

## Now, answer the following questions using the eq_passengercars dataset:

**1. Which country has the highest "Trade value of exports"? Which year was this?**

```{r}
highest_trade_exports = eq_passengercars_tidy %>%
# Filter the rows that correspond to "Trade value of exports"
  filter(measurement == "Trade value of exports") %>%
# filter the row where "value" is the maximum
  filter(value == max(value, na.rm = TRUE)) %>%
  select(country, year, value)
```

```{r}
# Germany had the highest Trade value of exports in 2014 
highest_trade_exports
```

**2. Which year had the lowest "Trade value of exports"? Which country was this?**

```{r}
lowest_trade_exports = eq_passengercars_tidy %>%
# Filter the rows that correspond to "Trade value of exports"
  filter(measurement == "Trade value of exports") %>%
# filter the row where "value" is the maximum
  filter(value == min(value, na.rm = TRUE)) %>%
  select(country, year, value)
```

```{r}
# South Africa had the highest Trade value of exports in 1964 
highest_trade_exports
```

**3. Which year did your country/any country of your interest have the highest "Export Quality Index"?**

```{r}
highest_export_index = eq_passengercars_tidy %>%
# Filter the rows that correspond to "Export Quality Index"
  filter(measurement == "Export Quality Index") %>%
# I'll be choosing the United States  
  filter(country == "United States") %>%
# filter the row where "value" is the maximum
  filter(value == max(value, na.rm = TRUE)) %>%
  select(country, year, value)
```

```{r}
# The United States had the highest Export Quality Index in 1984 
highest_export_index
```

**4. Which year, on average, had the greatest decrease in "Trade value of exports" from the previous year?**

```{r}
# Summarize the dataset: filter for "Trade value of exports" and calculate the average export value per year.
trade_value_exports = eq_passengercars_tidy %>%
  filter(measurement == "Trade value of exports") %>%
  group_by(year) %>%
  summarise(avg_trade_exports = mean(value, na.rm = TRUE)) %>%
  arrange(year)
```

```{r}
# Create a new column representing the previous year.
year_prev = trade_value_exports %>%
  mutate(prev_year = year - 1)
```

```{r}
# Perform lef_join to match current year with previous' average
trade_comparison = year_prev %>%
  left_join(trade_summary, by = c("prev_year" = "year"), suffix = c("", "_prev"))
```

```{r}
# Calculate the decrease from the previous year 
decrease_year = trade_comparison %>%
  filter(!is.na(avg_trade_exports_prev)) %>%
  mutate(decrease = avg_trade_exports_prev - avg_trade_exports) %>%
  filter(decrease == max(decrease)) %>%
  select(year, decrease)
```

```{r}
# The year that had the greatest decrease in "Trade value of exports" was 2009
decrease_year
```

**5. Can you think of the reason this might have happened?**

There could be multiple reasons as to why there might have been a significant decrease. The first reason that comes to mind is the U.S economic crisis that led to a significant decrease in exports. As economies slow down, suppliers provide less supplies and consumers buy less products. Hence, leasing to a decrease in export activity. Moreover, the government might have changed trade policies by imposing tariffs, sanctions, and other trade barriers that can disrupt trade activity.  Meanwhile, there are varying other possibilities, such as geopolitical tensions and natural disasters that can hinder global supply chains and undermine the trade value of exports.

**6. Which country-year did the "Export Quality Index" have the highest increase from the previous year?**

```{r}
# Filter the dataset for "Export Quality Index" 
export_quality_index = eq_passengercars_tidy %>%
  filter(measurement == "Export Quality Index") %>%
  select(country, year, value) %>%
  mutate(prev_year = year - 1)
```

```{r}
export_comparison = export_quality_index %>%
  left_join(eqi_data, by = c("country" = "country", "prev_year" = "year"), 
            suffix = c("", "_prev"))
```

```{r}
# Calculate the increase from the previous year
export_comparison <- export_comparison %>%
  mutate(increase = value - value_prev)
```

```{r}
# Calculate the highest increase
country_year_export = eqi_comparison %>%
  filter(!is.na(increase)) %>%
  filter(increase == max(increase, na.rm = TRUE)) %>%
  select(country, year, increase)
```

```{r}
# The country with the highest increase was Congo in 1966
country_year_export
```

# Question 2 - Mapping

For this question, I decided to use data for the **Employment Status for the Population 16 Years and Over** for 3 states in the U.S **(Illinois, Indiana, & Florida)**
